Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 code-tests.tex
test problems through the Test Orbit (11.2.7)

The general structure of each test problem description is as follows.
We will describe the construction of the test problem (including its
initial and boundary conditions), the analytical or expected
solutions, and motivate why we have included it in the paper.  After
that, we will show and describe \enzo's solution to the test problem.

>>>>

The general structure of each test problem description is as follows.
We will describe the construction of the test problem (including its
initial and boundary conditions), the analytical or expected
solutions, and motivate why we have included it in the paper.  After
that, we will show and describe \enzo's solution to the test problem.
We remind the user that, as discussed at the end of
Section~\ref{sec.tests.suite}, they can download a Mercurial
repository containing this paper, the \enzo\ test problem parameter
files that produced the simulation data used to generate the figures
in this section, and the \texttt{yt} scripts necessary to create the
figures themselves.

### 2013-06-25 tests-orbit.tex
test problems through the Test Orbit (11.2.7)

\begin{figure}
\begin{center}
\includegraphics[width=0.45\textwidth]{figures/TestOrbit_xy.eps}
\includegraphics[width=0.45\textwidth]{figures/TestOrbit_TotalEnergy.eps}
\caption{Test particle behavior in the Orbit problem, which is used to
test the gravity solver and particle integration.  An effectively
massless test particle is set in circular orbit around a massive
central particle, with the gravitational potential calculated on a 3D,
$32^3$ grid, and evolved forward in time for 200 orbits.  Left panel:
path of the test particle in the orbital plane plotted for the
duration of the simulation.  Right panel: total specific energy
(kinetic plus potential) for the test particle for the duration of the
simulation.  The particle's mean energy (integrated over the length of
the simulation) is represented by a horizontal dashed line, and +/-
one standard deviation by horizontal dotted lines.  All quantities are
in arbitrary units.}
\label{fig.orbittest}
\end{center}
\end{figure}

The ``orbit test'' simulates the orbit of an effectively massless
particle around a massive primary particle and is used to test the
accuracy of both the gravity solver and particle integration.  In this
problem (shown in Figure~\ref{fig.orbittest}), a 3D box with unit size
and $32^3$ grid is created, and a particle with mass 1 is placed at
the center of the box (both box length and particle mass are in
arbitrary units).  A second ``test'' particle, with mass of $10^{-6}$,
is placed at a distance of 0.3 L$_{box}$ from the primary particle in
the x-y plane (at $z=0.5$~L$_{box}$).  Both particles are given
velocities such that they will orbit their common center of mass in
the x-y plane.  The simulation is run for 200 orbits, writing
out a dataset 
approximately once per orbit (with the particle positions being
written out several hundred times per orbit).

The left panel of Figure~\ref{fig.orbittest} shows the path of the
test particle in the x-y plane (at z=$0.5$), plotted over the duration
of the simulation.  The test particle
maintains a path that has effectively unnoticeable deviation from
circularity, and with no discernable drift in either net radius or
orbit center over time.  The right panel displays the total specific
energy of the test particle (defined as $\frac{1}{2} {\bf v}^2 +
\phi$, where $\phi$ is the gravitational potential at the position of
the test particle) as a function of time, plotted once per orbit.  The
mean total specific energy is $-1.76797$ in arbitrary units, with a
standard deviation of $0.004885$ ($0.276\%$ of the mean value), and
with a maximum deviation from the mean of $0.546\%$ of the mean value.
Given that the potential is calculated on $32^3$ cells, the two
particles are separated by between 10 and 14 cells depending on their
relative positioning in the grid, and the particles are integrated
using a second-order leapfrog method, this level of accuracy is
expected.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=0.45\textwidth]{figures/TestOrbit_xy.eps}
\includegraphics[width=0.45\textwidth]{figures/TestOrbit_TotalEnergy.eps}
\caption{Test particle behavior in the Orbit problem, which is used to
test the gravity solver and particle integration.  An effectively
massless test particle is set in circular orbit around a massive
central particle, with the gravitational potential calculated on a 3D,
$32^3$ grid, and evolved forward in time for 200 orbits.  Left panel:
path of the test particle in the orbital plane plotted for the
duration of the simulation, with a zoomed box showing a small portion
of the orbit to highlight deviations from perfect circularity.  Right panel: total specific energy
(kinetic plus potential) for the test particle for the duration of the
simulation.  The particle's mean energy (integrated over the length of
the simulation) is represented by a horizontal dashed line, and +/-
one standard deviation by horizontal dotted lines.  All quantities are
in arbitrary units.}
\label{fig.orbittest}
\end{center}
\end{figure}

The ``orbit test'' simulates the orbit of an effectively massless
particle around a massive primary particle and is used to test the
accuracy of both the gravity solver and particle integration.  In this
problem (shown in Figure~\ref{fig.orbittest}), a 3D box with unit size
and a $32^3$ grid is created, and a particle with mass 1 is placed at
the center of the box (both box length and particle mass are in
arbitrary units).  A second ``test'' particle, with mass of $10^{-6}$,
is placed at a distance of 0.3 L$_{box}$ from the primary particle in
the x-y plane (at $z=0.5$~L$_{box}$).  Both particles are given
velocities such that they will orbit their common center of mass in
the x-y plane.  The simulation is run for 200 orbits, writing
out a dataset 
approximately once per orbit (with the particle positions being
written out several hundred times per orbit).

The left panel of Figure~\ref{fig.orbittest} shows the path of the
test particle in the x-y plane (at z=$0.5$), plotted over the duration
of the simulation.  The test particle
maintains a path that has effectively unnoticeable deviation from
circularity, and with no discernable drift in either net radius or
orbit center over time.  The right panel displays the total specific
energy of the test particle (defined as $\frac{1}{2} {\bf v}^2 +
\phi$, where $\phi$ is the gravitational potential at the position of
the test particle) as a function of time, plotted once per orbit.  The
mean total specific energy is $-1.76797$ in arbitrary units, with a
standard deviation of $0.004885$ ($0.276\%$ of the mean value), and
with a maximum deviation from the mean of $0.546\%$ of the mean value.
Given that the potential is calculated on $32^3$ cells, the two
particles are separated by between 10 and 14 cells depending on their
relative positioning in the grid, and the particles are integrated
using a second-order leapfrog method, this level of accuracy is
expected and acceptable.

### 2013-06-25 tests-sedov.tex
test problems through the Test Orbit (11.2.7)

The Sedov Blast Test \citep{Sedov1959} models an intense explosion,
initiated by depositing thermal energy into a homogenous distribution
of gas. The result is a strong spherical shock wave centered on the
point of energy injection.  This problem is a popular test of
astrophysical simulation codes for three reasons.  First, this problem
is representative of the astrophysical phenomenon of a supernova
explosion.   Second, an analytical solution to the evolution of the
Sedov blast-wave exists, allowing for a direct test of the accuracy of
the code.  The radius of the shock front as a function of time is
given by

@@

In the initial state, the box contains a homogenous distribution of
gas at a density of 1 (note that, in the absense of gravity, this
problem is scale-free and thus without units). Thermal energy is
deposited into a single cell at the center of the box with $E_0 =
10.0$. The problem is run in two dimensions with reflecting
boundary conditions and a box size of side
$1$.  We selecte a top grid of $100 \times 100$
cells and a maximum of four levels of adaptive refinement, refining on
shocks and the slopes of the density and total energy fields. The
exception to this scheme is in the initial conditions, where
additional refinement is placed directly around the injection point to
better resolve spherical geometry. The results are assessed
at $t = 0.07$, which corresponds to a time just before the shock
reaches the box edge (see Figure~\ref{fig.sedov1}).

>>>>

The Sedov Blast test \citep{Sedov1959} models an intense explosion,
initiated by depositing thermal energy into a homogenous distribution
of gas. The result is a strong spherical shock wave centered on the
point of energy injection.  This problem is a popular test of
astrophysical simulation codes for three reasons.  First, this problem
is representative of the astrophysical phenomenon of a supernova
explosion.   Second, an analytical solution to the evolution of the
Sedov blast wave exists, allowing for a direct test of the accuracy of
the code.  The radius of the shock front as a function of time is
given by

@@

In the initial state, the box contains a homogenous distribution of
gas at a density of 1 (note that, in the absence of gravity or
radiative cooling, this
problem is scale-free and thus without units). Thermal energy is
deposited into a single cell at the center of the box with $E_0 =
10.0$. The problem is run in two dimensions with reflecting
boundary conditions and a box having a length of
$1$ in both directions.  We selecte a top grid of $100 \times 100$
cells and a maximum of four levels of adaptive refinement, refining by
factors of two on
shocks and the slopes of the density and total energy fields. The
exception to this scheme is in the initial conditions, where
additional refinement is placed directly around the injection point to
better resolve spherical geometry. The results are assessed
at $t = 0.07$, which corresponds to a time just before the shock
reaches the box edge (see Figure~\ref{fig.sedov1}).

### 2013-06-25 tests-shockpool.tex
test problems through the Test Orbit (11.2.7)

For Zeus (middle row of Figure~\ref{fig.shockpool}), the shock is broadened because of artificial viscosity and there are slightly more post-shock oscillations, although again quite mild.  The impact of entering and exiting the refined region is somewhat larger than in PPM; however, the most noticeable difference is the incorrect position of the shock front. This is due to the fact that the scheme is not energy-conserving (see also the Sedov problem, below), and has very little to do with the passage through the refined region.

Finally, the MUSCL scheme produces a shock which is intermediate in width between the two previous cases.  This method is energy conserving and so correctly reproduces the shock speed.  The oscillations are mild except for the cell immediately outside of the refined region upon exiting.

>>>>

For Zeus (middle row of Figure~\ref{fig.shockpool}), the shock is broadened because of artificial viscosity and there are slightly more post-shock oscillations, although again quite mild.  The impact of entering and exiting the refined region is somewhat larger than in PPM; however, the most noticeable difference is the incorrect position of the shock front. This is due to the fact that the scheme is not energy-conserving (see also the Sedov problem in Section~\ref{sec.tests.sedov}), and has very little to do with the passage through the refined region.

Finally, the MUSCL scheme produces a shock that is intermediate in width between the two previous cases.  This method is energy-conserving, and thus correctly reproduces the shock speed.  The oscillations are mild except for the cell immediately outside of the refined region upon exiting.

### 2013-06-25 tests-sodshocktube.tex
test problems through the Test Orbit (11.2.7)

We begin the paper test suite with the classic one-dimensional Sod shock tube problem
\citep{Sod78}, which provides a good test of a hydrodynamical solver's
ability to resolve a clean Riemann problem with clear separation
between the three resultant waves.  These waves consist of a rarefaction fan, a
contact discontinuity, and a moderate shock.  The initial state is 
$\rho_{\rm L}, P_{\rm L} = 1.0, 1.0$ on the left of the boundary at 
$x=0.5$ and $\rho_{\rm R}, P_{\rm R} = 0.125, 0.1$ on the right.
All velocities are initially zero.  In
Figure~\ref{fig.sodshocktube}, we show the density solution at the
final time ($t=0.25$) for three of our hydro solvers -- the spatially third-order
PPM, as well as the two second-order \zeus\ and MUSCL schemes.  We use
100 cells across the domain, (which is a relatively standard choice in
code method papers) and show solutions both with and without adaptive
mesh refinement.  Without AMR (top row in Figure \ref{fig.sodshocktube}), it is
clear that the PPM scheme produces by far the cleanest solution with
all wave families crisply reproduced (in particular, the contact
discontinuity and shock).  \zeus\ and MUSCL produce similar results,
with MUSCL doing a slightly better job on the rarefaction fan.  In Figure
\ref{fig.sodshocktube}, we also show the integrated absolute deviation from 
the exact solution, $||E_1|| = \sum_i \Delta x_i |F(x_i)-{\rm Exact}(x_i)|$.
%, as well as the maximum deviation,
%$||E_\infty|| = \mathrm{max_i}\left\{ \Delta x_i|F(x_i)-{\rm Exact}(x_i)| \right\}$.  
These numbers confirm the qualitative differences noted
previously.

>>>>

We begin the paper test suite with the classic one-dimensional Sod shock tube problem
\citep{Sod78}, which provides a good test of a hydrodynamical solver's
ability to resolve a clean Riemann problem with clear separation
between the three resultant waves.  These waves consist of a rarefaction fan, a
contact discontinuity, and a moderate shock.  The initial state is 
$\rho_{\rm L}, P_{\rm L} = 1.0, 1.0$ on the left of the boundary at 
$x=0.5$ and $\rho_{\rm R}, P_{\rm R} = 0.125, 0.1$ on the right.
All velocities are initially zero.  In
Figure~\ref{fig.sodshocktube}, we show the density solution at the
final time ($t=0.25$) for three of our hydro solvers -- the spatially third-order
PPM, as well as the two second-order \zeus\ and MUSCL schemes.  We use
100 cells across the domain, which is a relatively standard choice in
code method papers, and show solutions both with and without adaptive
mesh refinement.  Without AMR (top row in Figure \ref{fig.sodshocktube}), it is
clear that the PPM scheme produces by far the cleanest solution with
all wave families crisply reproduced (in particular, the contact
discontinuity and shock).  \zeus\ and MUSCL produce similar results,
with MUSCL doing a slightly better job on the rarefaction fan.  In Figure
\ref{fig.sodshocktube}, we also show the integrated absolute deviation from 
the exact solution, $||E_1|| = \sum_i \Delta x_i |F(x_i)-{\rm Exact}(x_i)|$.
%, as well as the maximum deviation,
%$||E_\infty|| = \mathrm{max_i}\left\{ \Delta x_i|F(x_i)-{\rm Exact}(x_i)| \right\}$.  
These numbers confirm the qualitative differences noted
previously.

### 2013-06-25 tests-wavepool.tex
test problems through the Test Orbit (11.2.7)

In this one-dimensional test we pass a short wavelength linear wave through a static singley-refined region.  The full domain is from 0 to 1 and the refined region is set from 0.25 to 0.75 at all times.  We modify the left boundary consistent with a single linear sound wave with density given by $\rho(x,t) = \rho_0 (1 + A \sin(kx - \omega t))$, where the amplitude is $A = 0.01$ and the wavelength $\lambda = k/2\pi = 0.1$, and similar expressions exist for the pressure and velocity.  The initial, unperturbed density and pressure are set to unity and we adopt $\gamma = 1.4$.  The (unrefined) root grid is covered with 100 cells, resulting in a wavelength for the linear wave of only 10 cells.  This is, therefore, a challenging problem for hydro methods.  We are particularly interested in any reflection or artifacts introduced by the wave entering and exiting the refined regions.

@@

For the Zeus solver, we see that even before it has entered the refined region there are small oscillations excited behind the wave; although it is also worth noting that the wave itself is beautifully propagated without significant damping or phase errors.  In this test we use only our standard, low amount of quadratic artificial viscosity.  These waves could be damped by additional viscosity, but we do not add any in order to be sensitive to any artifacts at the grid boundary.  The remaining panels show that the trailing oscillations continue, but do not generate any additional noise. We note that the end result is similar to the case without any refined region.

>>>>

In this one-dimensional test we pass a short wavelength linear wave through a static singly-refined region.  The full domain is from 0 to 1 and the refined region is set from 0.25 to 0.75 at all times.  We modify the left boundary consistent with a single linear sound wave with density given by $\rho(x,t) = \rho_0 (1 + A \sin(kx - \omega t))$, where the amplitude is $A = 0.01$ and the wavelength $\lambda = k/2\pi = 0.1$.  Similar expressions exist for the pressure and velocity.  The initial, unperturbed density and pressure are set to unity and we adopt $\gamma = 1.4$.  The (unrefined) root grid is covered with 100 cells, resulting in a wavelength for the linear wave of only 10 cells.  This is, therefore, a challenging problem for hydro methods.  We are particularly interested in any reflection or artifacts introduced by the wave entering and exiting the refined regions.

@@

For the Zeus solver, we see that even before it has entered the refined region there are small oscillations excited behind the wave, although it is also worth noting that the wave itself is beautifully propagated without significant damping or phase errors.  In this test we use only our standard, low amount of quadratic artificial viscosity.  These waves could be damped by additional viscosity, but we do not add any in order to be sensitive to any artifacts at the grid boundary.  The remaining panels show that the trailing oscillations continue, but do not generate any additional noise. We note that the end result is similar to the case without any refined region.
